---
title: "Joseph Kohl Report - Jkohl9770@floridapoly.edu"
output: html_document
date: "2026-06-06"
---

```{r}
library(tidyverse)
```

```{r}
sleep <- read.csv("../data/SleepStudy.csv")
summary(sleep)
```
```{r}
#Looking at average sleep distribution to ensure it looks normal
ggplot(sleep,aes(x=AverageSleep)) +
  geom_histogram(binwidth = .3,fill="blue",color="black") +
  labs(
    title="Average Sleep Distribution",
    x="Average Sleep (hours)",
    y="Number of Students"
  ) + theme_classic()
```

This chart displays how many hours of sleep students get on average.

The data is normally distributed (bell-shaped) and peaks around 7 to 8 hours of sleep per night. Very few students average fewer than 5 hours or more than 10 hours of sleep.

```{r}
#Looking at avg sleep vs gpa
library(plotly)
gpa_interactive <- ggplot(sleep, aes(x = AverageSleep, y = GPA)) +
  geom_point(aes(color = factor(ClassYear)), size = 2) +
  geom_smooth(aes(color = factor(ClassYear)), method = "loess", se = FALSE) +
  geom_smooth(aes(color = "Overall"), method = "loess", se = FALSE, linetype = "dashed") +
  scale_color_manual(
    values = c("1" = "#E69F00", "2" = "#56B4E9", "3" = "#009E73", "4" = "#D55E00", "Overall" = "#000000")
  ) +
  labs(
    title = "Average Sleep vs GPA",
    color = "Class Year",
    x = "Average Sleep (Hours)",
    y = "GPA"
  ) + 
  theme_minimal()

ggplotly(gpa_interactive)
```

This plot examines whether sleep duration correlates with cumulative GPA, with students grouped by their class year (Freshman through Senior).

There is a general upward trend approaching 7-8 hours, but after that more sleep seems to become a negative rather than a positive.

```{r}
#Looking into effect of sleep behavior on gpa
ggplot(sleep, aes(x=LarkOwl,y=GPA,fill=LarkOwl)) +
  geom_boxplot() +
  labs(
    title="GPA vs Sleep Type",
    x="Sleep Type: Lark (Early Bird), Owl (Night Owl), Neither",
    y="Student GPA"
  ) + theme_minimal()
```

This chart compares the GPAs of students based on their self-identified sleep profile: Larks (early birds), Owls (night owls), or "Neither".

Students who identify as "Larks" have a slightly higher median GPA than "Owls." Students who identify as "Neither" sit right in the middle, indicating that waking up early is a slight advantage but not entirely definitive as seen by the outliers in Lark.

```{r}
#Scatter Plots showing Sleep Quality vs Amount affect on depression
ggplot(sleep, aes(x=PoorSleepQuality,y=DepressionScore)) +
  geom_point() + geom_smooth(method = "lm",se=FALSE) +
  labs(
    title = "Sleep Quality vs Depression Score",
    x="Poor Sleep Quality Measurement",
    y="Depression Score"
  ) + theme_minimal()


ggplot(sleep, aes(x=AverageSleep,y=DepressionScore)) +
  geom_point() + geom_smooth(method = "lm",se=FALSE) +
  labs(
    title = "Average Sleep Amount vs Depression Score",
    x="Average Sleep (Hours)",
    y="Depression Score"
  ) + theme_minimal()
```

These two  plots evaluate how sleep relates to mental health.

Sleep Quality: There is a clear positive correlation; as poor sleep quality measurements increase, depression scores also rise.

Sleep Amount: There is a little to no correlation between sleep time and depression, indicating that quality sleep is far more important than total time.

```{r}
#Drinking rates vs happiness

ggplot(sleep, aes(x=Drinks,y=Happiness))+
  geom_point(aes(color=AlcoholUse)) +
  geom_smooth(color="Black",se=FALSE) +
  labs(
    Title="Drinking Rates versus Happiness",
    color="Alcohol Use Category",
    x="Drinks Per Week",
    y="Happiness Score"
  ) + theme_minimal()
```

This plot looks at the relationship between weekly alcohol consumption (number of drinks) and self-reported happiness scores.

Happiness scores experience a very slight peak at moderate levels of alcohol consumption around 7 drinks per week. However, once consumption exceeds roughly 10 drinks per week, happiness levels drop off sharply.

## Conclusion

When looking at the dataset as a whole, a clear picture emerges regarding the relationship between student lifestyle choices, academic success, and mental well-being:

Both sleep duration (how long you sleep) and sleep timing (when you sleep) play a role in academic success. Students who get more sleep and those who align with an early-bird ("Lark") schedule consistently achieve higher GPAs.

The data strongly supports the idea that sleep deprivation and poor sleep quality are risk factors for depressive symptoms. Maximizing quality sleep hours is correlated with lower depression scores.

Lifestyle habits like drinking show non-linear relationships with happiness. While light-to-moderate social drinking does not show a negative relationship with happiness, high consumption is strongly linked to lower self-reported happiness.

For a student aiming to maximize both their GPA and mental well-being, the optimal lifestyle profile involves maintaining an early sleep schedule, averaging 7–9 hours of high-quality sleep per night, and keeping weekly alcohol consumption low to moderate.